---
title: "Decadal Trends in Seasonal Climatic Variables in Dar es Salaam, Tanzania: A Non-Parametric Approach Using the Mann-Kendall Test"
author: "January G. Msemakweli"
date: "2025-11-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Mann-Kendall Trend Analysis for Seasonal Climatic Variables

This R Markdown document reproduces the Mann–Kendall trend analysis for seasonal climatic variables in Dar es Salaam, Tanzania. The analysis uses monthly data from the Tanzania Meteorological Authority covering the period 2014–2024. The Mann–Kendall test is a non-parametric statistical test used to detect monotonic trends in time series data, making it particularly suitable for climatic data analysis as it does not require assumptions about data distribution. Sen's slope estimator is used in conjunction with the Mann–Kendall test to quantify the magnitude of detected trends.

### Methodology

The analysis follows these key steps:
1. **Data Preparation**: Load and format the climatic dataset with proper date handling
2. **Variable Selection**: Identify the climatic variables of interest (rainfall, temperatures, humidity)
3. **Trend Detection**: Apply the Mann–Kendall test to detect statistically significant trends
4. **Trend Quantification**: Calculate Sen's slope to estimate the rate of change
5. **Results Presentation**: Generate formatted tables summarizing the statistical findings

```{r}
library(readr)
library(dplyr)
library(Kendall)
library(trend)
library(knitr)
library(kableExtra)
library(ggplot2)
library(tidyr)

df <- read_csv("CLIMATE DAR DATASET.csv", show_col_types = FALSE)
df$date <- as.Date(paste(df$Year, df$Month, "01", sep = "-"), format="%Y-%b-%d")

vars <- c("rainfall_mm",
          "daytime_temperature_celcius",
          "nighttime_temperature_celcius",
          "relative_humidity_%")

var_labels <- c("Rainfall (mm)",
                "Daytime Temperature (°C)",
                "Nighttime Temperature (°C)",
                "Relative Humidity (%)")

run_mk <- function(x) {
  mk <- MannKendall(x)
  sen_result <- sens.slope(x)
  sen <- sen_result$estimates
  sen_ci <- sen_result$conf.int
  S <- mk$S
  varS <- mk$varS
  
  if (varS > 0) {
    if (S > 0) {
      Z <- (S - 1) / sqrt(varS)
    } else if (S == 0) {
      Z <- 0
    } else {
      Z <- (S + 1) / sqrt(varS)
    }
  } else {
    Z <- 0
  }
  
  return(list(Z = Z, p = mk$sl, slope = sen, ci_lower = sen_ci[1], ci_upper = sen_ci[2]))
}

results <- data.frame()

for (s in unique(df$Season)) {
  sub_data <- df %>% filter(Season == s) %>% arrange(date)
  
  for (i in seq_along(vars)) {
    v <- vars[i]
    label <- var_labels[i]
    x <- sub_data[[v]]
    
    if (length(x) >= 8) {
      mkout <- run_mk(x)
      results <- rbind(results, data.frame(
        Season = s,
        Variable = label,
        Z = round(mkout$Z, 4),
        p_value = round(mkout$p, 5),
        Sen_slope = round(mkout$slope, 5),
        CI_lower = round(mkout$ci_lower, 5),
        CI_upper = round(mkout$ci_upper, 5)
      ))
    }
  }
}

# Calculate Bonferroni correction
n_tests <- nrow(results)
results$p_value_bonferroni <- pmin(results$p_value * n_tests, 1)

results %>%
  mutate(
    Sen_slope_CI = paste0(round(Sen_slope, 5), " [", round(CI_lower, 5), ", ", round(CI_upper, 5), "]"),
    p_value_corrected = round(p_value_bonferroni, 5)
  ) %>%
  select(Season, Variable, Z, p_value, p_value_corrected, Sen_slope_CI) %>%
  kable(
    format = "html",
    caption = "Mann–Kendall Trend Test Results for Seasonal Climatic Variables (2014–2024)",
    col.names = c("Season", "Variable", "Z-Score", "p-value", "p-value (Bonferroni)", "Sen's slope [95% CI]"),
    digits = 5
  ) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

```

## Statistical Power Analysis: Minimum Detectable Trends

This section assesses the statistical power of the trend analysis by calculating the Minimum Detectable Trend (MDT) at 80% power and estimating the approximate power for each observed Sen's slope. This helps contextualize non-significant findings—distinguishing between genuine stability and insufficient power to detect real trends.

```{r}
library(dplyr)
library(knitr)
library(kableExtra)

power_slope_ols <- function(x, alpha = 0.05, power_target = 0.80){
  x <- x[!is.na(x)]
  n <- length(x)
  if(n < 8) return(NULL)

  t <- 1:n
  fit <- lm(x ~ t)
  sigma <- summary(fit)$sigma
  Sxx <- sum((t - mean(t))^2)

  se_b <- sigma / sqrt(Sxx)

  # Approximate MDT for target power (two-sided)
  tcrit <- qt(1 - alpha/2, df = n - 2)
  tpow  <- qt(power_target, df = n - 2)
  mdt <- (tcrit + tpow) * se_b

  list(n=n, sigma=sigma, se_b=se_b, MDT=mdt)
}

power_for_given_slope <- function(x, slope, alpha = 0.05){
  x <- x[!is.na(x)]
  n <- length(x)
  if(n < 8) return(NA_real_)

  t <- 1:n
  fit <- lm(x ~ t)
  sigma <- summary(fit)$sigma
  Sxx <- sum((t - mean(t))^2)
  se_b <- sigma / sqrt(Sxx)

  tcrit <- qt(1 - alpha/2, df = n - 2)
  ncp <- abs(slope) / se_b

  # Normal approximation to power
  pnorm(ncp - tcrit) + pnorm(-ncp - tcrit)
}

power_table <- data.frame()

for (s in unique(df$Season)) {
  sub_data <- df %>% filter(Season == s) %>% arrange(date)

  for (i in seq_along(vars)) {
    v <- vars[i]
    label <- var_labels[i]
    x <- sub_data[[v]]

    out <- power_slope_ols(x, alpha = 0.05, power_target = 0.80)
    if(!is.null(out)){

      obs <- results %>% filter(Season == s, Variable == label)
      obs_sen <- if(nrow(obs) > 0) as.numeric(obs$Sen_slope[1]) else NA_real_
      obs_power <- if(!is.na(obs_sen)) power_for_given_slope(x, obs_sen, alpha = 0.05) else NA_real_

      power_table <- rbind(power_table, data.frame(
        Season = s,
        Variable = label,
        n = out$n,
        Residual_SD = out$sigma,
        MDT_80pct = out$MDT,
        Observed_SenSlope = obs_sen,
        Power_for_Observed = obs_power
      ))
    }
  }
}

power_table_fmt <- power_table %>%
  mutate(
    Residual_SD = round(Residual_SD, 3),
    MDT_80pct = round(MDT_80pct, 4),
    Observed_SenSlope = round(Observed_SenSlope, 4),
    Power_for_Observed = round(Power_for_Observed, 2)
  )

power_table_fmt %>%
  kable(
    format = "html",
    caption = "Minimum Detectable Trend (MDT) at 80% power and approximate power for observed Sen slopes (OLS-based approximation; α=0.05).",
    col.names = c("Season", "Variable", "n", "Residual SD", "MDT (80% power)", "Observed Sen slope", "Power for observed slope")
  ) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  footnote(
    general = "MDT is the smallest trend magnitude expected to be detected with ~80% power at α=0.05 under an OLS approximation. Power values are approximate and are used to contextualize non-significant findings (e.g., relative humidity) as potentially underpowered rather than confirming stability.",
    general_title = "Note:"
  )
```

**Interpretation of Power Analysis Results:**

The power analysis above provides critical context for interpreting Mann–Kendall test results. Non-significant findings (p ≥ 0.05) should not be automatically interpreted as evidence of no trend—they may instead reflect insufficient statistical power to detect existing trends, particularly when the observed slope is smaller than the Minimum Detectable Trend (MDT). Variables with power estimates below 0.80 for their observed slopes are considered potentially underpowered. Additionally, statistically significant trends with observed slopes near the MDT threshold represent relatively weak signals that, while detectable, may have limited practical significance. Notably, **Relative Humidity (RH)** tends to exhibit high residual variability, which inflates the MDT and reduces power; thus, non-significant RH trends should be interpreted with caution as potentially reflecting power limitations rather than genuine climatic stability. This power assessment helps distinguish between true absence of trends and analytical limitations due to sample size and variability.

## Lag-1 Autocorrelation Diagnostics

```{r}
# Calculate lag-1 autocorrelation for each time series
autocorr_results <- data.frame()

for (s in unique(df$Season)) {
  sub_data <- df %>% filter(Season == s) %>% arrange(date)
  
  for (i in seq_along(vars)) {
    v <- vars[i]
    label <- var_labels[i]
    x <- sub_data[[v]]
    
    if (length(x) >= 8) {
      # Remove NA values for autocorrelation calculation
      x_clean <- x[!is.na(x)]
      if (length(x_clean) >= 8) {
        # Calculate lag-1 autocorrelation
        acf_result <- acf(x_clean, lag.max = 1, plot = FALSE)
        lag1_acf <- acf_result$acf[2]
        
        # Determine implications for Type I error
        # Positive autocorrelation inflates Type I error
        # Negative autocorrelation deflates Type I error
        if (abs(lag1_acf) < 0.1) {
          implication <- "Negligible effect on Type I error"
          risk_level <- "Low"
        } else if (lag1_acf > 0.1 & lag1_acf < 0.3) {
          implication <- "Slight inflation of Type I error"
          risk_level <- "Moderate"
        } else if (lag1_acf >= 0.3) {
          implication <- "Substantial inflation of Type I error (may require pre-whitening)"
          risk_level <- "High"
        } else if (lag1_acf < -0.1 & lag1_acf > -0.3) {
          implication <- "Slight deflation of Type I error"
          risk_level <- "Moderate"
        } else {
          implication <- "Substantial deflation of Type I error"
          risk_level <- "High"
        }
        
        autocorr_results <- rbind(autocorr_results, data.frame(
          Season = s,
          Variable = label,
          Lag1_ACF = round(lag1_acf, 4),
          Implication = implication,
          Risk_Level = risk_level
        ))
      }
    }
  }
}

autocorr_results %>%
  kable(
    format = "html",
    caption = "Lag-1 Autocorrelation Diagnostics and Implications for Type I Error",
    col.names = c("Season", "Variable", "Lag-1 ACF", "Implication for Type I Error", "Risk Level"),
    digits = 4
  ) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  footnote(
    general = "Lag-1 autocorrelation (ACF) measures the correlation between consecutive observations. Positive autocorrelation can inflate Type I error rates in trend tests, leading to false positives. Values |ACF| < 0.1 indicate negligible effect; 0.1 ≤ |ACF| < 0.3 indicate moderate effect; |ACF| ≥ 0.3 indicate substantial effect and may require pre-whitening procedures.",
    general_title = "Note:"
  )

```

## Trend-Free Pre-Whitening (TFPW) Analysis

```{r}
# Trend-Free Pre-Whitening function
# Based on Wang & Swail (2001) and Yue et al. (2002)
tfpw <- function(x) {
  # Remove NA values
  x_clean <- x[!is.na(x)]
  n <- length(x_clean)
  
  if (n < 8) {
    return(NA)
  }
  
  # Step 1: Estimate trend using Sen's slope
  sen_slope <- sens.slope(x_clean)$estimates
  t <- 1:n
  
  # Step 2: Remove trend
  x_detrended <- x_clean - sen_slope * t
  
  # Step 3: Estimate AR(1) coefficient from detrended series
  if (n > 1) {
    # Calculate lag-1 autocorrelation
    acf_result <- acf(x_detrended, lag.max = 1, plot = FALSE)
    rho <- acf_result$acf[2]
    
    # Constrain rho to [-0.99, 0.99] to avoid numerical issues
    rho <- max(-0.99, min(0.99, rho))
    
    # Step 4: Pre-whiten the detrended series
    x_pw <- numeric(n)
    x_pw[1] <- x_detrended[1]
    for (i in 2:n) {
      x_pw[i] <- x_detrended[i] - rho * x_detrended[i-1]
    }
    
    # Step 5: Add trend back
    x_tfpw <- x_pw + sen_slope * t
    
    return(x_tfpw)
  } else {
    return(NA)
  }
}

# Re-run analysis with TFPW for series with |ACF| >= 0.3
comparison_results <- data.frame()

for (s in unique(df$Season)) {
  sub_data <- df %>% filter(Season == s) %>% arrange(date)
  
  for (i in seq_along(vars)) {
    v <- vars[i]
    label <- var_labels[i]
    x <- sub_data[[v]]
    
    if (length(x) >= 8) {
      # Remove NA values
      x_clean <- x[!is.na(x)]
      if (length(x_clean) >= 8) {
        # Calculate lag-1 ACF
        acf_result <- acf(x_clean, lag.max = 1, plot = FALSE)
        lag1_acf <- acf_result$acf[2]
        
        # Original MK test results
        mkout_orig <- run_mk(x_clean)
        
        # Determine if pre-whitening is needed
        needs_pw <- abs(lag1_acf) >= 0.3
        
        if (needs_pw) {
          # Apply TFPW
          x_tfpw <- tfpw(x_clean)
          
          if (is.numeric(x_tfpw) && length(x_tfpw) >= 8 && !any(is.na(x_tfpw))) {
            # Re-run MK test on pre-whitened series
            mkout_pw <- run_mk(x_tfpw)
            
            # Determine if inference changed
            # Inference changed if: significance status changed OR direction changed
            orig_sig <- mkout_orig$p < 0.05
            pw_sig <- mkout_pw$p < 0.05
            orig_dir <- sign(mkout_orig$slope)
            pw_dir <- sign(mkout_pw$slope)
            
            inference_changed <- (orig_sig != pw_sig) | (orig_dir != pw_dir)
            
            comparison_results <- rbind(comparison_results, data.frame(
              Season = s,
              Variable = label,
              Lag1_ACF = round(lag1_acf, 4),
              Pre_whitened = "Yes",
              Z_original = round(mkout_orig$Z, 4),
              p_original = round(mkout_orig$p, 5),
              slope_original = round(mkout_orig$slope, 5),
              Z_pw = round(mkout_pw$Z, 4),
              p_pw = round(mkout_pw$p, 5),
              slope_pw = round(mkout_pw$slope, 5),
              Inference_Changed = ifelse(inference_changed, "Yes", "No")
            ))
          } else {
            # TFPW failed, use original results
            comparison_results <- rbind(comparison_results, data.frame(
              Season = s,
              Variable = label,
              Lag1_ACF = round(lag1_acf, 4),
              Pre_whitened = "Yes (failed)",
              Z_original = round(mkout_orig$Z, 4),
              p_original = round(mkout_orig$p, 5),
              slope_original = round(mkout_orig$slope, 5),
              Z_pw = NA,
              p_pw = NA,
              slope_pw = NA,
              Inference_Changed = "N/A"
            ))
          }
        } else {
          # No pre-whitening needed
          comparison_results <- rbind(comparison_results, data.frame(
            Season = s,
            Variable = label,
            Lag1_ACF = round(lag1_acf, 4),
            Pre_whitened = "No",
            Z_original = round(mkout_orig$Z, 4),
            p_original = round(mkout_orig$p, 5),
            slope_original = round(mkout_orig$slope, 5),
            Z_pw = round(mkout_orig$Z, 4),
            p_pw = round(mkout_orig$p, 5),
            slope_pw = round(mkout_orig$slope, 5),
            Inference_Changed = "N/A"
          ))
        }
      }
    }
  }
}

# Format comparison table
comparison_results %>%
  mutate(
    Original_MK = paste0("Z=", Z_original, ", p=", p_original),
    PW_MK = ifelse(Pre_whitened == "No", 
                   paste0("Z=", Z_pw, ", p=", p_pw),
                   ifelse(is.na(Z_pw), "N/A", paste0("Z=", Z_pw, ", p=", p_pw))),
    Slope_Change = ifelse(Pre_whitened == "No",
                          paste0(slope_original, " (unchanged)"),
                          ifelse(is.na(slope_pw), "N/A", 
                                 paste0(slope_original, " → ", slope_pw)))
  ) %>%
  select(Season, Variable, Lag1_ACF, Pre_whitened, Original_MK, PW_MK, 
         Slope_Change, Inference_Changed) %>%
  kable(
    format = "html",
    caption = "Comparison of Original and Pre-Whitened Mann-Kendall Test Results",
    col.names = c("Season", "Variable", "Lag-1 ACF", "Pre-whitened?", 
                  "Original MK (Z, p)", "Pre-whitened MK (Z, p)", 
                  "Sen's Slope Change", "Inference Changed?"),
    digits = 5
  ) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  footnote(
    general = "Trend-Free Pre-Whitening (TFPW) was applied only to series with |lag-1 ACF| ≥ 0.3. Inference changed indicates whether statistical significance (p < 0.05) or trend direction changed after pre-whitening. Series with |ACF| < 0.3 were not pre-whitened and show identical results in both columns.",
    general_title = "Note:"
  )

```

## Pettitt Change-Point Test

The Pettitt test is a non-parametric test used to detect a single change-point in a time series. It identifies the location where a significant shift in the central tendency of the data occurs, which is particularly useful for identifying abrupt changes in climatic variables.

```{r}
# Pettitt Change-Point Test
pettitt_results <- data.frame()

for (s in unique(df$Season)) {
  sub_data <- df %>% filter(Season == s) %>% arrange(date)
  
  for (i in seq_along(vars)) {
    v <- vars[i]
    label <- var_labels[i]
    x <- sub_data[[v]]
    
    if (length(x) >= 8) {
      # Remove NA values
      x_clean <- x[!is.na(x)]
      dates_clean <- sub_data$date[!is.na(x)]
      
      if (length(x_clean) >= 8) {
        # Perform Pettitt test
        pettitt_out <- pettitt.test(x_clean)
        
        # Get change point location
        cp_index <- pettitt_out$estimate
        cp_date <- dates_clean[cp_index]
        cp_year <- format(cp_date, "%Y")
        
        # Calculate means before and after change point
        mean_before <- mean(x_clean[1:cp_index], na.rm = TRUE)
        mean_after <- mean(x_clean[(cp_index + 1):length(x_clean)], na.rm = TRUE)
        change_magnitude <- mean_after - mean_before
        change_direction <- ifelse(change_magnitude > 0, "Increase", "Decrease")
        
        pettitt_results <- rbind(pettitt_results, data.frame(
          Season = s,
          Variable = label,
          K_statistic = pettitt_out$statistic,
          p_value = round(pettitt_out$p.value, 5),
          Change_Point_Index = cp_index,
          Change_Point_Year = cp_year,
          Mean_Before = round(mean_before, 3),
          Mean_After = round(mean_after, 3),
          Change_Magnitude = round(change_magnitude, 3),
          Change_Direction = change_direction
        ))
      }
    }
  }
}

# Display Pettitt test results
pettitt_results %>%
  mutate(
    Significance = ifelse(p_value < 0.05, "Significant", "Not Significant"),
    Mean_Change = paste0(Mean_Before, " → ", Mean_After, " (", 
                         ifelse(Change_Magnitude > 0, "+", ""), Change_Magnitude, ")")
  ) %>%
  select(Season, Variable, K_statistic, p_value, Change_Point_Year, Mean_Change, Significance) %>%
  kable(
    format = "html",
    caption = "Pettitt Change-Point Test Results for Seasonal Climatic Variables (2014–2024)",
    col.names = c("Season", "Variable", "K Statistic", "p-value", "Change Point Year", 
                  "Mean Before → After (Change)", "Significance"),
    digits = 5
  ) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  footnote(
    general = "The Pettitt test detects a single change-point in the time series. The K statistic indicates the strength of the change, and the change point year shows when the shift occurred. Mean Before and After show the average values before and after the detected change point.",
    general_title = "Note:"
  )

```

## Panel Plot of Significant Trends

```{r}
df_long <- df %>%
  select(date, Season, all_of(vars)) %>%
  pivot_longer(cols = all_of(vars), names_to = "var", values_to = "value") %>%
  mutate(Variable = var_labels[match(var, vars)])

sig_results <- results %>%
  filter(p_value < 0.05) %>%
  select(Season, Variable) %>%
  distinct()

df_long_sig <- df_long %>%
  inner_join(sig_results, by = c("Season", "Variable"))

df_panel <- df_long_sig %>%
  group_by(Season, Variable) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(t = row_number()) %>%
  group_modify(~ {
    sen <- sens.slope(.x$value)$estimates
    intercept <- median(.x$value - sen * .x$t, na.rm = TRUE)
    .x %>%
      mutate(
        sen_slope = sen,
        intercept = intercept,
        sen_line = intercept + sen * t
      )
  }) %>%
  ungroup()

p1 <- ggplot(df_panel, aes(x = date)) +
  geom_point(aes(y = value)) +
  geom_line(aes(y = sen_line), linewidth = 1) +
  facet_grid(Variable ~ Season, scales = "free_y") +
  labs(x = "Year", y = "Value")

print(p1)
ggsave("panel_significant_trends.png", p1, width = 10, height = 8, dpi = 300)
```

### Complete Time Series for All Variables and Seasons

```{r}
df_all_panel <- df_long %>%
  group_by(Season, Variable) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(t = row_number()) %>%
  group_modify(~ {
    if (length(.x$value) >= 8) {
      sen <- sens.slope(.x$value)$estimates
      intercept <- median(.x$value - sen * .x$t, na.rm = TRUE)
      .x %>%
        mutate(
          sen_slope = sen,
          intercept = intercept,
          sen_line = intercept + sen * t
        )
    } else {
      .x %>%
        mutate(
          sen_slope = NA,
          intercept = NA,
          sen_line = NA
        )
    }
  }) %>%
  ungroup() %>%
  left_join(results, by = c("Season", "Variable")) %>%
  mutate(Significance = if_else(p_value < 0.05, "Significant (p < 0.05)", "Not Significant"))

p2 <- ggplot(df_all_panel, aes(x = date, y = value)) +
  geom_point(aes(color = Significance), alpha = 0.6, size = 1) +
  geom_line(aes(y = sen_line, linetype = Significance), linewidth = 1.2, na.rm = TRUE) +
  scale_color_manual(values = c("Significant (p < 0.05)" = "#d62728", 
                                 "Not Significant" = "#636363")) +
  scale_linetype_manual(values = c("Significant (p < 0.05)" = "solid", 
                                   "Not Significant" = "dashed")) +
  facet_grid(Variable ~ Season, scales = "free_y") +
  labs(
    x = "Year",
    y = "Value",
    color = "Trend Significance",
    linetype = "Trend Significance",
    subtitle = "Solid lines indicate statistically significant trends (p < 0.05); dashed lines indicate non-significant trends"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    strip.text = element_text(size = 10),
    legend.position = "bottom"
  )

print(p2)
ggsave("complete_time_series.png", p2, width = 12, height = 10, dpi = 300)
```

### Seasonal Box Plots

```{r}
p3 <- ggplot(df_long, aes(x = Season, y = value)) +
  stat_boxplot(geom = "errorbar", width = 0.6, linewidth = 0.7, color = "black", 
               coef = 1.5) +
  geom_boxplot(fill = "white", color = "black", outlier.color = "black", 
               linewidth = 0.7, fatten = 1, coef = 1.5, 
               width = 0.6, outlier.size = 1.5, show.legend = FALSE) +
  facet_wrap(~ Variable, scales = "free_y", ncol = 2) +
  labs(x = "Season", y = "Value") +
  theme_bw() +
  theme(
    panel.grid.major = element_line(color = "white", linewidth = 0.5),
    panel.grid.minor = element_line(color = "white", linewidth = 0.25),
    panel.background = element_rect(fill = "lightgrey"),
    panel.border = element_rect(color = "black", fill = NA),
    strip.background = element_rect(fill = "grey", color = "black"),
    strip.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(color = "black"),
    legend.position = "none"
  )

print(p3)
ggsave("seasonal_boxplots.png", p3, width = 10, height = 7, dpi = 300)
```

### Sen's Slope Magnitude Visualization

```{r}
results_viz <- results %>%
  mutate(
    Significance = if_else(p_value < 0.05, "Significant", "Not Significant"),
    Trend_Direction = if_else(Sen_slope > 0, "Increasing", "Decreasing")
  )

p4 <- ggplot(results_viz, aes(x = reorder(paste(Season, Variable, sep = " - "), Sen_slope), 
                        y = Sen_slope, 
                        fill = Significance)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.5) +
  coord_flip() +
  scale_fill_manual(values = c("Significant" = "#d62728", 
                                "Not Significant" = "#95a5a6")) +
  labs(
    x = "Season - Variable",
    y = "Sen's Slope (units/year)",
    fill = "Statistical Significance"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.text.y = element_text(size = 8),
    legend.position = "bottom"
  )

print(p4)
ggsave("sens_slope_magnitude.png", p4, width = 10, height = 8, dpi = 300)
```

### Annual Trends in Climatic Variables by Season (2014-2024)
#### Dashed lines show linear trend fits with 95% confidence intervals

```{r}
df_annual <- df %>%
  group_by(Year, Season) %>%
  summarise(
    Rainfall = mean(rainfall_mm, na.rm = TRUE),
    Daytime_Temp = mean(daytime_temperature_celcius, na.rm = TRUE),
    Nighttime_Temp = mean(nighttime_temperature_celcius, na.rm = TRUE),
    Humidity = mean(`relative_humidity_%`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Rainfall, Daytime_Temp, Nighttime_Temp, Humidity),
    names_to = "Variable",
    values_to = "Annual_Mean"
  ) %>%
  mutate(
    Variable = case_when(
      Variable == "Rainfall" ~ "Rainfall (mm)",
      Variable == "Daytime_Temp" ~ "Daytime Temperature (°C)",
      Variable == "Nighttime_Temp" ~ "Nighttime Temperature (°C)",
      Variable == "Humidity" ~ "Relative Humidity (%)"
    )
  )

p5 <- ggplot(df_annual, aes(x = Year, y = Annual_Mean, color = Season)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_line(linewidth = 1, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, linetype = "dashed", linewidth = 1, alpha = 0.3) +
  facet_wrap(~ Variable, scales = "free_y", ncol = 2) +
  labs(
    x = "Year",
    y = "Annual Mean Value",
    color = "Season"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    strip.text = element_text(size = 10),
    legend.position = "bottom"
  ) +
  scale_color_brewer(palette = "Set2")

print(p5)
ggsave("annual_trends.png", p5, width = 10, height = 8, dpi = 300)
```

### Monthly Climatology of Climatic Variables (2014-2024)
#### Average monthly values showing typical annual cycles

```{r}
df_monthly <- df %>%
  mutate(Month_Num = as.numeric(format(date, "%m"))) %>%
  group_by(Month_Num, Season) %>%
  summarise(
    Rainfall = mean(rainfall_mm, na.rm = TRUE),
    Daytime_Temp = mean(daytime_temperature_celcius, na.rm = TRUE),
    Nighttime_Temp = mean(nighttime_temperature_celcius, na.rm = TRUE),
    Humidity = mean(`relative_humidity_%`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Rainfall, Daytime_Temp, Nighttime_Temp, Humidity),
    names_to = "Variable",
    values_to = "Monthly_Mean"
  ) %>%
  mutate(
    Variable = case_when(
      Variable == "Rainfall" ~ "Rainfall (mm)",
      Variable == "Daytime_Temp" ~ "Daytime Temperature (°C)",
      Variable == "Nighttime_Temp" ~ "Nighttime Temperature (°C)",
      Variable == "Humidity" ~ "Relative Humidity (%)"
    ),
    Month_Name = month.abb[Month_Num]
  ) %>%
  mutate(Month_Name = factor(Month_Name, levels = month.abb))

p6 <- ggplot(df_monthly, aes(x = Month_Name, y = Monthly_Mean, color = Season, group = Season)) +
  geom_point(size = 2.5) +
  geom_line(linewidth = 1) +
  facet_wrap(~ Variable, scales = "free_y", ncol = 2) +
  labs(
    x = "Month",
    y = "Average Value",
    color = "Season"
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_line(color = "white", linewidth = 0.5),
    panel.grid.minor = element_line(color = "white", linewidth = 0.25),
    panel.background = element_rect(fill = "lightgrey"),
    panel.border = element_rect(color = "black", fill = NA),
    strip.background = element_rect(fill = "grey", color = "black"),
    strip.text = element_text(color = "black", size = 10),
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(color = "black")
  ) +
  scale_color_brewer(palette = "Set2")

print(p6)
ggsave("monthly_climatology.png", p6, width = 10, height = 6, dpi = 300)
```

### Mann-Kendall Z-Scores by Season and Variable
#### Horizontal dashed lines indicate critical values (±1.96) for α = 0.05

```{r}
results_viz2 <- results %>%
  mutate(
    Significance = if_else(p_value < 0.05, "Significant", "Not Significant"),
    Z_abs = abs(Z)
  )

p7 <- ggplot(results_viz2, aes(x = reorder(paste(Season, Variable, sep = " - "), Z), 
                         y = Z, 
                         fill = Significance)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_hline(yintercept = c(-1.96, 0, 1.96), 
             linetype = c("dashed", "solid", "dashed"), 
             color = c("red", "black", "red"), 
             linewidth = 0.7) +
  annotate("text", x = 1, 
           y = c(-1.96, 1.96), 
           label = c("Z = -1.96\n(α = 0.05)", "Z = 1.96\n(α = 0.05)"), 
           vjust = c(1.2, -0.2), hjust = 0, size = 3, color = "red") +
  coord_flip() +
  scale_fill_manual(values = c("Significant" = "#d62728", 
                               "Not Significant" = "#95a5a6")) +
  labs(
    x = "Season - Variable",
    y = "Z-Score",
    fill = "Statistical Significance"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.text.y = element_text(size = 8),
    legend.position = "bottom"
  )

print(p7)
ggsave("zscore_visualization.png", p7, width = 10, height = 8, dpi = 300)
```

### Trend Magnitudes by Variable Across Seasons
#### Positive values indicate increasing trends; negative values indicate decreasing trends

```{r}
p8 <- ggplot(results_viz, aes(x = Season, y = Sen_slope, fill = Trend_Direction)) +
  geom_bar(stat = "identity", alpha = 0.8, position = "dodge") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.5) +
  facet_wrap(~ Variable, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("Increasing" = "#d62728", 
                               "Decreasing" = "#2ca02c")) +
  labs(
    x = "Season",
    y = "Sen's Slope (units/year)",
    fill = "Trend Direction"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    strip.text = element_text(size = 10),
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

print(p8)
ggsave("trend_magnitudes_by_variable.png", p8, width = 10, height = 8, dpi = 300)
```
